Towards Practical First-Order Model Counting

### 2025-06-01 doc/paper/paper.tex
Remove arxiv version (for now) and update conference info

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

@@

\keywords{first-order model counting, knowledge compilation, lifted inference}

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

@@

\acknowledgements{The first two authors contributed equally. Part of the
  research was conducted while all authors were at the National University of
  Singapore.}

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{siunitx}
\usepackage{tikz}
\usepackage{microtype}
\usepackage[backgroundcolor=lightgray]{todonotes} % TODO: temporary
\usepackage{mathtools}

@@

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
sentence into equations that include the necessary base cases. To begin, we use
\textsc{Crane} to compile the sentence into the three components: $\mathcal{E}$,
$\mathcal{F}$, and $\mathcal{D}$ (as described in \cref{sec:compilation}). After
some algebraic simplifications (described below), $\mathcal{E}$ is passed to the
\FindBaseCases procedure (see \cref{sec:identifying}). For each base case
$f(\mathbf{x})$, we retrieve the sentence $\mathcal{F}(f)$ associated with the
function name $f$ and simplify it using the \Propagate procedure (explained in
detail in \cref{sec:propagating}). We do this by iterating over all indices of
$\mathbf{x}$, where $x_{i}$ is a constant, and using \Propagate to simplify
$\psi$ by assuming that domain $\mathcal{D}(f, i)$ has size $x_{i}$. Finally, on
\autoref{line:final}, \CompileWithBaseCases recurses on these simplified
sentences and adds the resulting base case equations to $\mathcal{E}$.

\paragraph*{Simplify}
The main responsibility of the \Simplify procedure is to handle the algebraic
pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here: $n$ is a variable,
$a, b \in \mathbb{N}_{0}$ are constants, and $f$ is an expression that may
depend on $m$. \Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$.

\begin{example}\label{example:bijectionstwo}
  Let us return to the bijection-counting problem from \cref{example:bijections}
  and its initial solution described in \cref{example:solution}. \Simplify
  transforms $g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$
  into $g(l, m) = g(l-1, m) + mg(l-1, m-1)$. Then \FindBaseCases identifies two
  base cases: $g(0, m)$ and $g(l, 0)$. In both cases, \CompileWithBaseCases
  recurses on the sentence $\mathcal{F}(g)$ simplified by assuming that one of
  the domains is empty. In the first case, we recurse on the sentence
  $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$, where $S$ is a predicate
  introduced by preprocessing with weights $w^{+}(S) = 1$ and $w^{-}(S) = -1$.
  Hence, we obtain the base case $g(0, m) = 0^{m}$. In the case of $g(l, 0)$,
  \Propagate{$\psi$, $\Gamma$, $0$} returns an empty sentence, resulting in
  $g(l, 0) = 1$. While these base cases overlap when $l = m = 0$, they remain
  consistent since $0^{0} = 1$.
\end{example}

% Generally, let $\phi$ be a sentence with two domains $\Gamma$ and $\Delta$,
% and let $n, m \in \mathbb{N}_{0}$. Then the FOMC of \Propagate{$\phi$,
% $\Delta$, $n$} assuming $|\Gamma| = m$ is the same as the FOMC of
% \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.

% Although \CompileWithBaseCases starts with a call to \textsc{Crane}, the
% proposed algorithm is not just a post-processing step for FOKC because
% \cref{alg:compilewithbasecases} is recursive and can issue more calls to
% \textsc{Crane} on various derived sentences.

We end this section by bounding the recursion depth of the \CompileWithBaseCases
procedure, thus proving that the algorithm terminates.

\begin{theorem}
  Let $\phi$ be a sentence with $n$ domains. Then the maximum recursion depth of
  \CompileWithBaseCases{$\phi$} is $n$.
\end{theorem}

To prove the theorem, we rely on two observations about the algorithms presented
in \cref{sec:identifying,sec:propagating}.

\begin{observation}\label{observation1}
  Each base case returned by \FindBaseCases has at least one constant (in line
  with \cref{def:basecase}).
\end{observation}

@@

\begin{proof}
  We proceed by induction on $n$. If $n=0$, then $\phi$ is essentially a
  propositional formula, and \textsc{Crane} compiles it into an equation of the
  form $f = \expr$ with no `function calls'. Suppose that---for all sentences
  with at most $n$ domains---\CompileWithBaseCases terminates with a recursion
  depth of at most $n$. Let $\phi$ be a sentence with $n+1$ domains. By
  \cref{observation1}, each base case on \autoref{line:loopy} of
  \cref{alg:compilewithbasecases} has at least one constant. Therefore, by
  \cref{observation2}, after \autoref{line:propagate}, sentence $\psi$ has at
  most $n$ domains. Thus, \autoref{line:final} terminates with a recursion depth
  of at most $n$ by the inductive hypothesis, completing the inductive proof.
\end{proof}

@@

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{function call $f(\mathbf{y})$ on the RHS of an equation in $\mathcal{E}$}{
    $\mathbf{x} \gets \text{the parameters of $f$ in its definition}$\;
    \ForEach{$y_{i} \in \mathbf{y}$} {
      \lIf{$y_{i} \in \mathbb{N}_{0}$}{$\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto y_{i}] \,\}$}
      \ElseIf{$y_{i} = x_{i} - c_{i}$ for some $c_{i} \in \mathbb{N}_{0}$}{
        \lFor{$j \gets 0$ \KwTo $c_{i} - 1$}{$\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto j] \,\}$}
      }
    }
  }
\end{algorithm}
% (later, let's not bother yet): replace x_i-c_i on Line 7 of Algorithm 2 with a
% generalisation of x_i that can be an arbitrary summation/subtraction of domain
% sizes (propagating this change to the rest of the paper, particularly
% Assumption 1, the proof of Lemma 1)

\Cref{alg:findbasecases} summarises the implementation of \FindBaseCases. It
considers two types of arguments when a function $f$ calls itself recursively:
constants and arguments of the form $x_{i} - c_{i}$. Here, $c_{i}$ is a
constant, and $x_{i}$ is the $i$-th argument of the signature of $f$. When the
argument is a constant $c_{i}$, a base case with $c_{i}$ is added. In the second
case, a base case is added for each constant from $0$ up to (but not including)
$c_{i}$.

@@

It can be shown that the base cases identified by \FindBaseCases are sufficient
for the algorithm to terminate.\footnote{Note that characterising the
  fine-grained complexity of the solutions found by \Cranetwo{} or other FOMC
  algorithms is an emerging area of research. These questions have been
  partially addressed in previous
  work~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/kr/TothK24}
  and are orthogonal to the goals of this section.} For the remainder of this
section, let $\mathcal{E}$ denote the equations returned by
\CompileWithBaseCases.

\begin{theorem}\label{thm:halting}
  Let $f$ be an $n$-ary function in $\mathcal{E}$ and
  $\mathbf{x} \in \mathbb{N}_{0}^{n}$. Then the evaluation of $f(\mathbf{x})$
  terminates.
\end{theorem}

We prove \cref{thm:halting} using double induction. First, we apply induction to
the number of functions in $\mathcal{E}$. Then, we use induction on the arity of
the `last' function in $\mathcal{E}$ according to some topological ordering. We
begin with a few observations that stem from
previous~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/ijcai/BroeckTMDR11} and this
work.

@@

\begin{observation}\label{assumption2}
  There is a \emph{topological ordering} of all functions ${(f_{i})}_{i}$ in
  $\mathcal{E}$ such that equations in $\mathcal{E}$ with $f_{i}$ on the LHS do
  not contain function calls to $f_{j}$ with $j > i$. This condition prevents
  mutual recursion and other cyclic scenarios.
\end{observation}

@@

\begin{assumption}\label{assumption4}
  For each equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every recursive
  function call $f(\mathbf{y}) \in \expr$ satisfies the following:
  \begin{itemize}
    \item Each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some constant
          $c_{i}$.
    \item There exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
          $c_{i} > 0$.
  \end{itemize}
\end{assumption}

Finally, we assume a particular order of evaluation for function calls using the
equations in $\mathcal{E}$. Specifically, we assume that base cases are
considered before the recursive definition. The exact order in which base cases
are considered is immaterial.

\begin{assumption}
  When multiple equations in $\mathcal{E}$ match a function call
  $f(\mathbf{x})$, preference is given to an equation with the most constants on
  its LHS.
\end{assumption}

@@

\begin{lemma}\label{lemma:oneunary}
  Assume that $\mathcal{E}$ consists of just \emph{one unary} function $f$. Then
  the evaluation of a function call $f(x)$ terminates for any
  $x \in \mathbb{N}_{0}$.
\end{lemma}
\begin{proof}
  If $f(x)$ is captured by a base case, then its evaluation terminates by
  \cref{fact,fact2}. If $f$ is not recursive, the evaluation of
  $f(x)$ terminates by \cref{fact}.

  Otherwise, let $f(y)$ be an arbitrary function call on the RHS of the
  definition of $f(x)$. If $y$ is a constant, then there is a base case for
  $f(y)$. Otherwise, let $y = x - c$ for some $c > 0$. Then there exists
  $k \in \mathbb{N}_{0}$ such that $0 \le x - kc \le c-1$. So, after $k$
  iterations, the sequence of function calls $f(x), f(x-c), f(x-2c),\dots$ will
  be captured by the base case $f(x \mod c)$.
\end{proof}

\begin{lemma}\label{lemma:onefunction}
  Generalising \cref{lemma:oneunary}, let $\mathcal{E}$ be a set of equations
  for \emph{one} $n$-ary function $f$ for some $n \ge 1$. Then the evaluation of
  $f(\mathbf{x})$ terminates for any $\mathbf{x} \in \mathbb{N}_{0}^{n}$.
\end{lemma}
\begin{proof}
  If $f$ is non-recursive, the evaluation of $f(\mathbf{x})$ terminates by
  previous arguments. We proceed by induction on $n$, with the base case of
  $n=1$ handled by \cref{lemma:oneunary}. Assume that $n > 1$. Any base case of
  $f$ can be seen as a function of arity $n-1$, since one of the parameters is
  fixed. Thus, the evaluation of any base case terminates by the inductive
  hypothesis. It remains to show that the evaluation of the recursive equation
  for $f$ terminates, but that follows from \cref{assumption3}.
\end{proof}

\begin{proof}[Proof of \cref{thm:halting}]
  We proceed by induction on the number of functions $n$. The base case of $n=1$
  is handled by \cref{lemma:onefunction}. Let ${(f_{i})}_{i=1}^{n}$ be some
  topological ordering of these $n > 1$ functions. If $f = f_{j}$ for $j < n$,
  then the evaluation of $f(\mathbf{x})$ terminates by the inductive hypothesis
  since $f_{j}$ cannot call $f_{n}$ by \cref{assumption2}. Using the inductive
  hypothesis that all function calls to $f_{j}$ (with $j < n$) terminate, the
  proof proceeds similarly to the Proof of \cref{lemma:onefunction}.
\end{proof}

@@

\Cref{alg:propagate}, called \Propagate, modifies the sentence $\phi$ based on
the assumption that $|\Delta| = n$. When $n=0$, some clauses become vacuously
satisfied and can be removed. When $n > 0$, partial grounding is performed by
replacing all variables quantified over $\Delta$ with constants. (None of the
sentences examined in this work had $n > 1$.) \Cref{alg:propagate} handles these
two cases separately. For a literal or a clause $C$, the set of corresponding
domains is denoted as $\Doms(C)$.

% NOTE: keep this enumerate
In the case of $n = 0$, there are three types of clauses to consider:
\begin{enumerate}
  \item those that do not mention $\Delta$,\label[type]{type1}
  \item those in which every literal contains variables quantified over
  $\Delta$, and\label[type]{type2}
  \item those that have some literals with variables quantified over $\Delta$
  and some without.\label[type]{type3}
\end{enumerate}
Clauses of \cref{type1} are transferred to the new sentence $\phi'$ without any
changes. For clauses of \cref{type2}, $C'$ is empty, so these clauses are
filtered out. As for clauses of \cref{type3}, a new kind of smoothing is
performed, which will be explained in \cref{sec:smoothing}.

@@

% TODO: maybe turn the footnote into a remark?

\begin{example}
  Let $C \equiv \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{.
  } \neg P(x, y) \lor \neg P(x, z) \lor y=z$. Then
  $\Doms(C) = \Doms(\neg P(x, y)) = \Doms(\neg P(x, z)) = \{\, \Gamma, \Delta \,\}$,
  and $\Doms(y=z) = \{\, \Delta \,\}$. A call to \Propagate{$\{\, C \,\}$,
    $\Delta$, $3$} would result in the following sentence with nine clauses:
  \begin{align*}
    (\forall x \in \Gamma\text{. }\neg P(x, c_{1}) \lor& \neg P(x, c_{1}) \lor c_{1}=c_{1})\land{}\\
    (\forall x \in \Gamma\text{. }\neg P(x, c_{1}) \lor& \neg P(x, c_{2}) \lor c_{1}=c_{2})\land{}\\
    \vdots&\\
    (\forall x \in \Gamma\text{. }\neg P(x, c_{3}) \lor& \neg P(x, c_{3}) \lor c_{3}=c_{3}).
  \end{align*}
  Here, $c_{1}$, $c_{2}$, and $c_{3}$ are the new constants.
\end{example}

@@

Smoothing modifies a circuit to reintroduce eliminated atoms, ensuring the
correct model count~\cite{darwiche2001tractable,DBLP:conf/ijcai/BroeckTMDR11}.
In this section, we describe a similar process performed on
lines~\ref{line:presmoothing}--\ref{line:smoothing2} of \cref{alg:propagate}.
Line~\ref{line:presmoothing} checks if smoothing is necessary, and
lines~\ref{line:smoothing1} and~\ref{line:smoothing2} execute it. If the
condition on \autoref{line:presmoothing} is not satisfied, the clause is not
smoothed but omitted.

Suppose \Propagate is called with arguments $(\phi, \Delta, 0)$, i.e., we are
simplifying the sentence $\phi$ by assuming that the domain $\Delta$ is empty.
Informally, if there is a predicate $P$ in $\phi$ unrelated to $\Delta$,
smoothing preserves all occurrences of $P$ even if all clauses with $P$ become
vacuously satisfied.

\begin{example}\label{example:basecasesmoothing}
  Let $\phi$ be
  \begin{align}
    (\forall x \in \Delta\text{. }\forall y, z \in \Gamma&\text{. }Q(x) \lor P(y, z))\land{}\label[clause]{eq:example1}\\
    (\forall y, z \in \Gamma'&\text{. }P(y, z))\label[clause]{eq:example2},
  \end{align}
  where $\Gamma' \subseteq \Gamma$ is a domain introduced by a compilation rule.
  It should be noted that $P$, as a relation, is a subset of
  $\Gamma \times \Gamma$.

@@

In this section, we will describe the final step of \Cranetwo{} as outlined in
\cref{fig:overview}, i.e., translating the set of equations $\mathcal{E}$ into
C++ code. Recall that this step is crucial for the usability of the algorithm,
otherwise function definitions would remain purely mathematical, with no
convenient way to compute the model count for particular domain sizes. Once a
C++ program is produced, it can be executed with different command-line
arguments to compute the model count of the sentence for various domain sizes.

>>>>

\documentclass[a4paper,UKenglish,cleveref,autoref,thm-restate]{lipics-v2021}
\pdfoutput=1
\bibliographystyle{plainurl}

@@

\keywords{First-order model counting, knowledge compilation, lifted inference}

\relatedversion{}

@@

\acknowledgements{The first two authors contributed equally. Part of the
  research was conducted while all authors were at the National University of
  Singapore.}
\nolinenumbers

\EventEditors{Jeremias Berg and Jakob Nordstr\"{o}m}
\EventNoEds{2}
\EventLongTitle{28th International Conference on Theory and Applications of Satisfiability Testing (SAT 2025)}
\EventShortTitle{SAT 2025}
\EventAcronym{SAT}
\EventYear{2025}
\EventDate{August 12--15, 2025}
\EventLocation{Glasgow, Scotland}
\EventLogo{}
\SeriesVolume{341}
\ArticleNo{5}

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{siunitx}
\usepackage{tikz}
\usepackage{microtype}
\usepackage{mathtools}

@@

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
sentence into equations that include the necessary base cases. First, we use
\textsc{Crane} to compile the sentence into three components: $\mathcal{E}$,
$\mathcal{F}$, and $\mathcal{D}$ (as described in \cref{sec:compilation}). After
some algebraic simplifications (described below), the algorithm passes
$\mathcal{E}$ to the \FindBaseCases procedure (see \cref{sec:identifying}). For
each base case $f(\mathbf{x})$, we retrieve the sentence $\mathcal{F}(f)$
associated with the function name $f$ and simplify it using the \Propagate
procedure (explained in detail in \cref{sec:propagating}). We do this by
iterating over all indices of $\mathbf{x}$, where $x_{i}$ is a constant, and
using \Propagate to simplify $\psi$ by assuming that domain $\mathcal{D}(f, i)$
has size $x_{i}$. Finally, on \autoref{line:final}, \CompileWithBaseCases
recurses on these simplified sentences and adds the resulting base case
equations to $\mathcal{E}$.

\paragraph*{Simplify}
The main responsibility of the \Simplify procedure is to handle the algebraic
pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here, $n$ is a variable, $a$ and
$b$ are constants, and $f$ is an expression that may depend on $m$. \Simplify
transforms this pattern into $f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$.

\begin{example}\label{example:bijectionstwo}
  {\par\sloppy
  We return to the bijection-counting problem from \cref{example:bijections} and
  its initial solution described in \cref{example:solution}. \Simplify
  transforms
  \[
    g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)
  \]
  into
  \[
    g(l, m) = g(l-1, m) + mg(l-1, m-1).
  \]
  Then \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In
  both cases, \CompileWithBaseCases recurses on the sentence $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the sentence $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by preprocessing with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we obtain the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty sentence, resulting in $g(l, 0) = 1$. While these base cases
  overlap when $l = m = 0$, they remain consistent since $0^{0} = 1$.
  \par}
\end{example}

Generally, let $\phi$ be a sentence with two domains $\Gamma$ and $\Delta$, and
let $n, m \in \mathbb{N}_{0}$. Then the FOMC of \Propagate{$\phi$, $\Delta$,
  $n$}, assuming $|\Gamma| = m$, is the same as the FOMC of \Propagate{$\phi$,
  $\Gamma$, $m$}, assuming $|\Delta| = n$.

Although \CompileWithBaseCases starts with a call to \textsc{Crane}, the
proposed algorithm is not merely a post-processing step for FOKC because
\cref{alg:compilewithbasecases} is recursive and can issue additional calls to
\textsc{Crane} on various derived sentences. We conclude this section by
bounding the recursion depth of the \CompileWithBaseCases procedure, thereby
also proving that the algorithm terminates.

\begin{theorem}
  Let $\phi$ be a sentence with $n$ domains. The maximum recursion depth of
  \CompileWithBaseCases{$\phi$} is then $n$.
\end{theorem}

The proof of this theorem relies on two observations regarding the algorithms
presented in \cref{sec:identifying,sec:propagating}.

\begin{observation}\label{observation1}
  Each base case returned by \FindBaseCases contains at least one constant (in
  line with \cref{def:basecase}).
\end{observation}

@@

\begin{proof}
  We proceed by induction on $n$. If $n=0$, then $\phi$ is essentially a
  propositional formula, and \textsc{Crane} compiles it into an equation of the
  form $f = \expr$ with no `function calls'. Suppose that---for all sentences
  with at most $n$ domains---\CompileWithBaseCases terminates with a recursion
  depth of at most $n$. Let $\phi$ be a sentence with $n+1$ domains. By
  \cref{observation1}, each base case on \autoref{line:loopy} of
  \cref{alg:compilewithbasecases} has at least one constant. Therefore, by
  \cref{observation2}, after \autoref{line:propagate}, the sentence $\psi$ has
  at most $n$ domains. Thus, \autoref{line:final} terminates with a recursion
  depth of at most $n$ by the inductive hypothesis, completing the inductive
  proof.
\end{proof}

@@

  $\mathcal{B} \gets \emptyset$\;
  \ForEach{function call $f(\mathbf{y})$ on the RHS of an equation in $\mathcal{E}$}{
    $\mathbf{x} \gets \text{the parameters of $f$ in its definition}$\;
    \ForEach{$y_{i} \in \mathbf{y}$} {
      \lIf{$y_{i} \in \mathbb{N}_{0}$}{$\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto y_{i}] \,\}$}\label{line:constant}
      \ElseIf{$y_{i} = x_{i} - c_{i}$ for some $c_{i} \in \mathbb{N}_{0}$}{
        \lFor{$j \gets 0$ \KwTo $c_{i} - 1$}{$\mathcal{B} \gets \mathcal{B} \cup \{\, f(\mathbf{x})[x_{i} \mapsto j] \,\}$}\label{line:nonconstant}
      }
    }
  }
\end{algorithm}

\Cref{alg:findbasecases} summarises the implementation of \FindBaseCases. It
considers two types of arguments when a function $f$ calls itself recursively:
constants and arguments of the form $x_{i} - c_{i}$. When the argument is a
constant $c_{i}$, \autoref{line:constant} adds a base case with $c_{i}$ to the
set of all base cases $\mathcal{B}$. In the second case,
\autoref{line:nonconstant} adds a base case to $\mathcal{B}$ for each constant
from $0$ to (but not including) $c_{i}$.

@@

In the rest of this section, we will show that the base cases identified by
\FindBaseCases are sufficient for the algorithm to terminate.\footnote{Note that
  characterising the fine-grained complexity of the solutions found by
  \Cranetwo{} or other FOMC algorithms is an emerging area of research. These
  questions have been partially addressed in previous
  work~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/kr/TothK24} and are unrelated to
  the goals of this section.} Let $\mathcal{E}$ denote the equations returned by
\CompileWithBaseCases.

\begin{theorem}\label{thm:halting}
  Given an $n$-ary function $f$ in $\mathcal{E}$ and
  $\mathbf{x} \in \mathbb{N}_{0}^{n}$, the evaluation of $f(\mathbf{x})$
  terminates.
\end{theorem}

We prove \cref{thm:halting} using double induction. First, we apply induction to
the number of functions in $\mathcal{E}$. Then, we use induction on the arity of
the `last' function in $\mathcal{E}$ according to a topological ordering. Before
proving \cref{thm:halting}, we make a few observations about this and
previous~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/ijcai/BroeckTMDR11} work.

@@

\begin{observation}\label{assumption2}
  There is a \emph{topological ordering} ${(f_{i})}_{i}$ of all functions in
  $\mathcal{E}$ such that equations in $\mathcal{E}$ with $f_{i}$ on the LHS do
  not contain function calls to $f_{j}$ with $j > i$.
\end{observation}

\Cref{assumption2} prevents mutual recursion and other cyclic scenarios.

@@

\begin{assumption}\label{assumption4}
  For each equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every recursive
  function call $f(\mathbf{y}) \in \expr$ satisfies the following:
  \begin{itemize}
    \item each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some constant
          $c_{i}$ and
    \item there exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
          $c_{i} > 0$.
  \end{itemize}
\end{assumption}

Finally, we assume a particular order of evaluation for function calls using the
equations in $\mathcal{E}$; specifically, base cases precede the recursive
definition.

\begin{assumption}
  When multiple equations in $\mathcal{E}$ match a function call
  $f(\mathbf{x})$, preference is given to the equation with the most constants
  on its LHS.
\end{assumption}

@@

\begin{lemma}\label{lemma:oneunary}
  Assume that $\mathcal{E}$ consists of just \emph{one unary} function called
  $f$. Then the evaluation of $f(x)$ terminates for any $x \in \mathbb{N}_{0}$.
\end{lemma}
\begin{proof}
  If $f(x)$ matches a base case, then its evaluation terminates by
  \cref{fact,fact2}. If $f$ is not recursive, the evaluation of $f(x)$ also
  terminates by \cref{fact}.

  Otherwise, let $f(y)$ be an arbitrary function call on the RHS of the
  definition of $f(x)$. If $y$ is a constant, then there is a base case for
  $f(y)$. Otherwise, let $y = x - c$ for some $c > 0$. Then there exists
  $k \in \mathbb{N}_{0}$ such that $0 \le x - kc \le c-1$. So, after $k$
  iterations, the sequence of function calls $f(x), f(x-c), f(x-2c),\dots$ will
  match the base case $f(x \mod c)$.
\end{proof}

\begin{lemma}\label{lemma:onefunction}
  Generalising \cref{lemma:oneunary}, let $\mathcal{E}$ be a set of equations
  for \emph{one} $n$-ary function $f$ for some $n \ge 1$. The evaluation of
  $f(\mathbf{x})$ terminates for any $\mathbf{x} \in \mathbb{N}_{0}^{n}$.
\end{lemma}
\begin{proof}
  If $f$ is non-recursive, the evaluation of $f(\mathbf{x})$ terminates by
  previous arguments. We proceed by induction on $n$, with the base case of
  $n=1$ handled by \cref{lemma:oneunary}. Assume that $n > 1$. Any base case of
  $f$ can be seen as a function of arity $n-1$ since one of the parameters is
  constant. Thus, the evaluation of any base case terminates by the inductive
  hypothesis. It remains to show that the evaluation of the recursive equation
  for $f$ terminates, but this follows from \cref{assumption3}.
\end{proof}

\begin{proof}[Proof of \cref{thm:halting}]
  We proceed by induction on the number of functions $n$. The base case of $n=1$
  is handled by \cref{lemma:onefunction}. Let ${(f_{i})}_{i=1}^{n}$ be some
  topological ordering of these $n > 1$ functions. If $f = f_{j}$ for $j < n$,
  then the evaluation of $f(\mathbf{x})$ terminates by the inductive hypothesis
  since $f_{j}$ cannot call $f_{n}$ by \cref{assumption2}. Using the inductive
  hypothesis that all function calls to $f_{j}$ (with $j < n$) terminate, the
  proof proceeds similarly to the proof of \cref{lemma:onefunction}.
\end{proof}

@@

\Cref{alg:propagate}, called \Propagate, modifies the sentence $\phi$ based on
the assumption that $|\Delta| = n$. When $n=0$, some clauses become vacuously
satisfied and can be removed. When $n > 0$, partial grounding replaces all
variables with domain $\Delta$ with constants. (None of the sentences examined
in this work had $n > 1$.) \Cref{alg:propagate} handles these two cases
separately. For a literal or clause $C$, we write the set of corresponding
domains as $\Doms(C)$. In the case of $n = 0$, there are three types of clauses
to consider:
\begin{enumerate}
  \item those that do not mention $\Delta$,\label[type]{type1}
  \item those in which every literal contains variables quantified over
  $\Delta$ and\label[type]{type2}
  \item those with some literals containing such variables and some
        without.\label[type]{type3}
\end{enumerate}
We transfer clauses of \cref{type1} to the new sentence $\phi'$ without any
changes. For clauses of \cref{type2}, $C'$ is empty, so these clauses are
filtered out. As for clauses of \cref{type3},
lines~\ref{line:presmoothing}--\ref{line:smoothing2} perform a new kind of
smoothing, the explanation of which we defer to \cref{sec:smoothing}.

@@

\begin{example}
  Let $C \coloneqq \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{.
  } \neg P(x, y) \lor \neg P(x, z) \lor y=z$. Then
  $\Doms(C) = \Doms(\neg P(x, y)) = \Doms(\neg P(x, z)) = \{\, \Gamma, \Delta \,\}$,
  and $\Doms(y=z) = \{\, \Delta \,\}$. A call to \Propagate{$\{\, C \,\}$,
    $\Delta$, $3$} would result in the following sentence with nine clauses:
  \begin{align*}
    (\forall x \in \Gamma\text{. }\neg P(x, c_{1}) \lor& \neg P(x, c_{1}) \lor c_{1}=c_{1})\land{}\\
    (\forall x \in \Gamma\text{. }\neg P(x, c_{1}) \lor& \neg P(x, c_{2}) \lor c_{1}=c_{2})\land{}\\
    \vdots&\\
    (\forall x \in \Gamma\text{. }\neg P(x, c_{3}) \lor& \neg P(x, c_{3}) \lor c_{3}=c_{3}).
  \end{align*}
  Here, $c_{1}$, $c_{2}$, and $c_{3}$ are the new constants.
\end{example}

@@

Smoothing modifies a circuit to reintroduce eliminated atoms, ensuring the
correct model count~\cite{darwiche2001tractable,DBLP:conf/ijcai/BroeckTMDR11}.
This section describes a similar process performed on
lines~\ref{line:presmoothing}--\ref{line:smoothing2} of \cref{alg:propagate}.
Line~\ref{line:presmoothing} checks if smoothing is necessary, and
lines~\ref{line:smoothing1} and~\ref{line:smoothing2} execute it. If the
condition on \autoref{line:presmoothing} is not satisfied, the clause is not
smoothed but omitted.

Suppose \CompileWithBaseCases calls \Propagate with arguments
$(\phi, \Delta, 0)$, i.e., we are simplifying the sentence $\phi$ by assuming
that the domain $\Delta$ is empty. Informally, if there is a predicate $P$ in
$\phi$ unrelated to $\Delta$, smoothing preserves all occurrences of $P$, even
if all clauses with $P$ become vacuously satisfied.

\begin{example}\label{example:basecasesmoothing}
  Let $\phi$ be
  \begin{align}
    (\forall x \in \Delta\text{. }\forall y, z \in \Gamma&\text{. }Q(x) \lor P(y, z))\land{}\label[clause]{eq:example1}\\
    (\forall y, z \in \Gamma'&\text{. }P(y, z))\label[clause]{eq:example2},
  \end{align}
  where $\Gamma' \subseteq \Gamma$ is a domain introduced by a compilation rule.
  Note that $P$, as a relation, is a subset of $\Gamma \times \Gamma$.

@@

In this section, we will describe the final step of \Cranetwo{} as outlined in
\cref{fig:overview}, i.e., translating the set of equations $\mathcal{E}$ into
C++ code. Recall that this step is crucial for the usability of the algorithm;
otherwise, function definitions would remain purely mathematical, with no
convenient way to compute the model count for particular domain sizes. Once a
C++ program is produced, it can be executed with different command-line
arguments to determine the model count of the sentence for various domain sizes.
